\vspace{-0.5cm} XSGRB - The X-shooter GRB afterglow sample

### 2017-10-24 document/XSGRB_sample.tex
Merge branch 'master' into writing_branch
# Conflicts:
#	document/XSGRB_sample.tex

For a brief moment, the energy output of a $\gamma$-ray burst (GRB) can rival
that of the stellar light integrated over the entirety of the observable
universe \citep{Kumar2015}. Bright across the entire electromagnetic spectrum,
the following afterglow makes possible the exploration of the intervening space,
otherwise dark. Since the discovery of the first $\gamma$-ray counterparts
nearly two decades ago\citep{Costa1997a, VanParadijs1997, Djorgovski1997,
	Frail1997, Wijers1998}, significant advances in the understanding of the underlying
phenomenology and potential usage has been made.

Categorized on the basis of their duration and spectral hardness, two main classes of GRBs exist, long and short \citep{Kouveliotou1993}. Now thoroughly established, the connection between long GRBs and supernovae (SNe) \citep{Galama1998, Hjorth2003a, Stanek2003, Woosley2006c} and the association of the location of long GRBs with the most star-forming regions in their host galaxies \citep{Fruchter2006, Svensson2010a}, make a stellar origin a likely explanation for the progenitor system. The seeming metal aversion \citep{Graham2013, Kruhler2015, Perley2016b} in selection of host at low redshift along with theoretical single-progenitor models preferring more pristine starting conditions \citep{Woosley2006d}, paints a picture where the core of a rapidly spinning, massive star undergoes gravitational collapse, although a certain degree of binarity \citep{Fryer2005, Podsiadlowski2010} in the progenitor population is required to explain the host metallicity distribution to super solar values \citep{Niino2011}, see \citet{Levan2016} for recent review. The driving engine in short GRBs has been established as the merger of two compact objects \citep{Metzger2010} through the detection of an associated kilonova \citep{Tanvir2013a, Berger2013a}. With the advent of the first gravitational wave (GW) detection \cite{Abbott2016a}, the electromagnetic counterpart in the form of a short

GRBs as cosmological probes

Do GRBs trace star formation

Afterglow spectroscopy

Complete samples

Only after observing more than 12000 damped Lyman-$\alpha$ absorbers (DLAs)
towards about 10$^5$ QSOs have 5 systems with
$\log({N_\mathrm{HI}/\mathrm{cm^{-2}}}) > 22$ been identified
\citep{Noterdaeme2012b}. Long GRB afterglow spectra, by contrast, reveal such
systems in the majority of cases \citep{Jakobsson2006a, Fynbo2009}. Whereas DLAs
towards QSOs are mostly limited to $1.8\lesssim z \lesssim 5$ due to the
atmospheric  UV-cutoff and increasing Lyman-blanketing at increasing redshifts
\citep[e.g.,][]{Rafelski2014}, GRBs allow us to see into the hearts of
star-forming galaxies over the full history of cosmic star formation from $z
\approx 0$ to $z > 8$ \citep[e.g.,][]{Tanvir2009, Salvaterra2009,
	Jakobsson2012}. With afterglow spectroscopy (throughout the electromagnetic
spectrum from X-rays to the sub-mm) we can hence characterize the properties of
star-forming galaxies over cosmic history in terms of  redshifts, metallicities,
molecular contents, ISM temperatures, UV-flux densities.

Content of paper

%There are several unanswered but fundamental questions that must be addressed in
%order to exploit the full potential of GRBs as cosmological probes. More than
%50\% of the \textit{Swift} bursts with measured redshift are at $z > 2$, and
%5--10\% are expected to be above $z=5$ \citep{Salvaterra2007,
%	Salvaterra2012,Jakobsson2012, Perley2016a}. A high redshift completeness is
%crucial for our understanding of the link between the number density of GRBs per
%unit redshift and the global star-formation history of the Universe, as measured
%by other means \citep[UV, FIR, sub-mm, see][]{Robertson2012}. The detection of
%GRBs at $z > 6$ shows that GRBs have become competitive as a tool to identifying
%galaxies at the highest redshifts and unsurpassed in providing detailed
%abundance information via absorption line spectroscopy \citep{Tanvir2012,
%	McGuire2016}.
%
%\smallskip
%
%%The sample of short GRBs with redshifts remains too small and too
%%incomplete to really
%%constrain their redshift distribution (Guetta \& Piran 2006, A\&A, 453, 823;
%%Salvaterra et al.\ 2008, MNRAS, 388, 6) making it of extreme importance to
%%enlarge as far as possible the number of known redshifts for this class of
%%sources.
%
%From March 2005 to March 2016 there have been about 350 {\it Swift} bursts
%complying with our sample selection criteria, and
%about half of them have measured redshifts. Among the latter subset, the team
%proposing these observations has measured about two thirds of the redshifts,
%mainly with FORS1/2 and X-shooter \citep[see][]{Fynbo2009, Jakobsson2012, Kruhler2012}. Our
%current aim is to build a sample superior to our previous low-resolution survey
%\citep{Fynbo2009}, both in terms of quantity and  quality of
%the spectra. 
%
%Due to
%the wide wavelength range of X-shooter with the same observation cover molecular
%H$_2$ absorption near the atmospheric cut-off and all the strong emission lines
%from the host in the NIR arm \citep[e.g.,][]{Friis2015}. In
%general, the wide wavelength coverage ensures that we always have features on
%which to base a redshift measurement as long as the afterglow is brighter than
%about 23 mag in either the $R$- or $z$-band. Frequently, emission lines are also
%detected from the underlying host, which also provide further information such
%as SFR and metallicity (the top right panel in Fig.~1 shows an example). Only
%for 7 out of more than 70 secured spectra could we not measure a redshift. With
%the X-shooter survey we provide {\bf metallicity measurements} for about 30\%
%(Voigt-profile fits) of the $z>1.7$ events. So far we have measured
%metallicities for more than 20 GRB afterglows with X-shooter. With the wide
%wavelength coverage of X-shooter we can study important chemical species as
%Zn\,\textsc{ii}, Cr\,\textsc{ii} and $\alpha$ elements over a much wider
%redshift range than what is possible with other instruments. As an example, we
%have measured a metallicity of $0.1 Z_\odot$ for GRB\,100219A at $z=4.669$
%(Th{\"o}ne et al.\ 2013, MNRAS, 428, 3590), $0.02Z_\odot$ for GRB\,111008A at
%$z=4.991$ (Sparre et al.\ 2014, ApJ, 785, 150) and $0.05Z_\odot$ for the
%$z=5.9125$ GRB\,130606A (Hartoog et al.\ 2015, A\&A, 580, 139). Reconciling the
%abundance patterns of GRB absorbers, other types of absorbers, QSO DLAs in
%particular, and old stars in the Local Group is an important long-term goal (see
%also Sparre et al.\ 2014, ApJ, 785, 150). Metallicities are also measured from
%host emission lines (Kr{\"u}hler et al.\ 2015, A\&A, 581, A125). GRB
%spectroscopy also allows us to determine the dust content of their environments,
%both through analysis of the depletion pattern and through measurement of the
%associated extinction (Japelj et al.\ 2015, A\&A, 451, 2050). This allows us to
%quantify the dust-to-metals ratio and its evolution with redshift (e.g., De Cia
%et al.\ 2013, A\&A, 560, 88; Zafar \& Watson 2013, A\&A, 560, 26). %Moreover,
%% GRB
%%afterglows allow investigation of random foreground absorbers, complementing
%%QSO sightline studies (e.g., Schulze et
%%al.\ 2012, A\&A, 546, 20).
%
%%In Fynbo et al. (2009, ApJS, 185,
%%526), we compare the host-galaxy HI
%%column density and metallicity (most of which has been measured by our group)
%%with a sample of damped \lya\ systems (DLAs) along QSO sightlines. We show that
%%most GRB-DLAs have a much larger column than QSO-DLAs.  The difference in
%%column density and metallicity for GRB and QSO sightlines shows that the total
%%cross-section for environments like those of GRBs must be less than a few
%%percent of the total cross-section for QSO-DLAs, and suggests that GRB
%%sightlines probe regions with high gas density.
%
%%The issue of production and {\bf release of Lyman continuum radiation} is
%%currently not fully understood, but the evidence suggests that star-forming
%%galaxies are dominating, in particular at the highest redshifts (e.g., Bianchi
%%et al.\ 2001, A\&A, 376, 1; Chen et al.\ 2007, ApJ, 667, L25). Also here
%%X-shooter spectroscopy will allow major progress due to the excellent blue
%%sensitivity.
%
%%GRB afterglows can probe dust in high-redshift star-forming galaxies
%%including exploring the {\bf 2175~\AA{} extinction bump}
%%(El\'iasd\'ottir et al.\ 2009, ApJ, 697, 1725; Zafar et al.\ 2011, A\&A,
%%732, 143).
%%Due to the broad wavelength  coverage we can do
%%this more accurately and over a broader range of redshifts with X-shooter.
%
%\smallskip
%
%We will also determine the frequency and properties of {\bf molecular
%	absorption} towards GRB absorbers. Molecular gas is a key element to catalyze
%the process of star formation, but prior to our program H$_2$ had been detected
%just in two cases (tentatively in Fynbo et al.\ 2006, A\&A, 451, L47; securely
%in Prochaska et al.\ 2009, ApJ, 691, L27). With our X-shooter program we have
%found three more systems (Kr\"uhler et al.\ 2013, A\&A, 557, 18; D'Elia et al.\
%2014, A\&A, 564, 38; Friis et al.\, 2015, MNRAS, 451, 167). We are currently
%analysing more of our spectra for less obvious molecular absorption and we
%expect to find more (a dedicated sample paper is addressing this issue).
%
%\vspace{0.2cm} A natural question to ask is: {\bf how long should this work
%	continue?} Our view is that we need to keep observing the afterglows as long as
%we have \textit{Swift} in operation. Also note that the program is still
%producing many papers and provides data for many theses (Box 9 and 10).
%\textit{Swift} is currently funded until 2018, but is likely to get more
%extensions given its overwhelming success. As mentioned above GRBs allow us to
%probe star-forming galaxies that are almost impossible to study in other ways
%both in terms of redshifts, galaxy luminosity function, and regions within
%galaxies. After 7 periods we have secured seven spectra of $z>4$ GRBs, of which
%three were of sufficient quality to allow abundance measurements (Th\"one et
%al.\ 2013, Sparre et al.\ 2014, Hartoog et al.\ 2015). GRBs offer the only way
%to derive chemical abundances for the gas phase of central, actively
%star-forming regions of high-$z$ galaxies. The program also maintains a very
%high discovery potential where we occasionally find something completely
%unexpected that provides interesting clues to puzzles in other fields, e.g.
%extinction of type Ia supernovae (Fynbo et al.\ 2014, A\&A, 572, 12). Each of
%these spectra are like precious jewels – it is a type of observation that can
%never be repeated and a class of sightlines that can only be studied while we
%have operating GRB satellites.
%
%It is also worth adding that we have build up a rather unique team spread over
%Europe from Granada to Reykjavik, which by now has reached a point where the
%distribution of night shifts, the scientific exploitation of the data is
%efficient and where we are open to all new members who wish to participate. As
%mentioned all data are public immediately.
%
%{\bf For all of these reasons, we need to keep building up the sample of GRB
%	afterglow spectra now as we may have to wait many years before a mission like
%	\textit{Swift} becomes available again.}
%
%A significant proportion of GRBs lack a bright optical afterglow (``{\bf dark
%	bursts}'', e.g., Jakobsson et al.\ 2004, ApJ, 617, L21; Melandri et al.\ 2012,
%MNRAS, 421, 1265). Some of these are at the highest redshifts ($z > 6$) and
%their observer-frame optical emission is absorbed by the IGM. The majority,
%however, suffer from large dust obscuration (e.g., Perley et al. 2009, AJ, 138,
%1690; Greiner et al.\ 2011, A\&A, 526, 30). Identifying such GRBs is important
%for constraining the fraction of obscured star formation. In both cases, NIR
%emission is expected. X-shooter can adequately study these objects, provided
%that a NIR counterpart is timely identified, for which we have the dedicated
%HAWKI run D.
%
%The detection of {\bf absorption line variability} can reveal the burst
%influence on the surrounding medium and in turn the absorber distance from the
%burst and its metallicity (Vreeswijk et al.\ 2007, A\&A 468, 83; D'Elia et al.\
%2009, ApJ, 694, 332; Th\"one et al.\ 2011, MNRAS, 414, 479; De Cia et al.\ 2012,
%A\&A, 545, 64; Hartoog et al.\ 2013). {\bf Short GRBs} originate in a
%substantially different environment compared to long GRBs. Short GRBs may be
%related to the merging of compact binaries and the coalescence time can be long
%enough to allow the progenitor system to move far away from the star formation
%site (Belczynski et al.\ 2002, ApJ, 571, 147). Up to now, however, no spectrum
%with a sufficient signal-to-noise ratio of a short GRB afterglow has been
%secured. A knowledge of the redshift distribution of short bursts is of key
%importance for the next generation of gravitational wave experiments, as they
%are the likely EM counterparts to their primary targets.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Comparison to previous samples.}	
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

This should also include a discussion of dark bursts. 
Cuchiara. BAT6 (Salvaterra), TOUGH

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\subsection{X-ray column densities vs. optically derived ones (Remove section)}	
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%	
%
%\begin{figure}
%	\centerline{\includegraphics[width=10cm]{figures/OAXR_NH_sphere.pdf}}
%	\caption{Optically derived HI column densities vs. X-ray derived ones taken from \citet{Buchner2016}. The colors of the points reflect the redshift of the burst. Circles mark the column densities published in \citet{Fynbo2009} and squares mark the ones derived as part of the sample presented here. Overplot in blue is the 1:1 correspondence, expected if the conditions local to the GRBs are similar to the local ISM abundances. If the metallicity is lower at the GRB site, the point move upwards in the plot. }
%	\label{fig:OAXR}
%\end{figure}	
%	
%	
%	
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\subsection{Why is the HI column truncated? (Remove section)}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%What is the expected transmission for a high-column HI, where the hydrogen primarily will be in the form of H2? Number of bursts with detected H2 and their HI column density. Rahmati Schaye, Krongold Prochaska.

>>>>

GRBs are interesting as they provide the opportunity to derive new insights
into a very wide range of topics in astrophysics. In that sense GRB
research is very multidisciplinary. The {\it Swift} satellite \citep{Gehrels09}, 
which was launched in 2004, has made
it possible to harvest the great potential in using GRBs as probes of
galaxies in the intergalactic medium, which was already hinted at by
results from earlier missions.

A crucial aspect of the great success of the {\it Swift} mission has been
the extensive ground based follow-up observations of the afterglows and of
the host galaxies of the GRBs. This work has involved a large community
of researchers both inside and outside the official {\it Swift} team. This fruitful
collaboration has been greatly facilitated by the open data access policy
of the {\it Swift} mission.

In the beginning of the "{\it Swift} era" most of the follow-up spectroscopy
was secured using low-resolution spectrographs (typically
R=$\lambda/\Delta\lambda$$<$1000) \citep[e.g.][]{Fynbo2009}. For a handful of
very bright afterglows high-resolution (typically R$>$20000) spectra were
secured.

The X-shooter spectrograph \citep{Vernet2011} is the first of the 2nd generation
instruments at the ESO Very Large Telescope (VLT). It is a high-throughput 
echelle spectrograph
covering the full spectra range from the UV atmospheric cutoff to the near-infrared
(near-IR)
K-band with three spectral arms (UVB, VIS, NIR). It was designed very much with
transient source follow-up in mind as the fading luminosities of such sources 
makes in urgent to secure as extensive coverage as possible in the shortest possible 
time. At the same time the resolution was set to be in the range 4000--9000 in order
to be able to get a large useful spectra range between the many sky-background
emission lines in the red and near-IR.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion and conclusions}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

In this paper we have presented the results
of a dedicated effort over the years 2009 -- 2016 to use the X-shooter 
spectrograph on the ESO-VLT to secure spectroscopic observations of 
afterglows and host galaxies of GRBs detected by {\it Swift}. This 
work was initiated by a consortium using Guaranteed Time on X-shooter, 
but over the years the project continuued in open time and the team was
opened to include any researchers interested in contribution to this 
effort.

The sample we present here include spectroscopic observations of 86 
systems fulfilling our sample criteria of which 72 are afterglow spectra 
and 14 are observations of the underlying host galaxies. 
UPDATE THIS.
